Apparent preferences for cognitive effort fade when multiple forms of effort

### 2021-05-28 Manuscript.Rmd
Exp 1 revised

  There were three block types, in which handling times of 2, 10, or 14 seconds were paired with travel times of 14, 6, and 2 seconds, respectively. All combinations added up to 16 seconds, meaning that accepted trials took the same amount of time in all conditions, but rejected trials were shorter in environments with shorter travel times. Timing parameters were held constant within each 7-min block. Each block type was presented twice in pseudorandom order (total session length = 42 minutes). Reward amounts varied per trial (4, 8, or 20 cents with equal probability), with the constraint that every reward was presented twice every six trials. This prevented sequences from being dominated by a single amount during any window of time. Timing information was disclosed at the beginning of each block, and each trial's reward amount was displayed during a 2 s offer window before the trial began. Participants received training prior to the experimental session and were told about all possible timing conditions and reward amounts in order to avoid any biases that could arise through experience-dependent learning [@Dundon2020; @Garrett2020]. Finally, we encouraged participants to carefully evaluate their options by telling them that accepting all offers would not maximize their rewards.

@@

# Proportion of performance-based forced travels per subject and cost type
# move to 16.2.1.
propFails <- dataBtw %>% 
  filter(Cost != "Wait") %>% 
  group_by(SubjID, Cost) %>% 
  summarise(propFails = mean(rawChoice == 2), .groups = 'drop') %>%
  group_by(Cost) %>% 
  summarise(meanFT = mean(propFails),
            sdFT = sd(propFails), .groups = 'drop')
```

@@

  In this experiment, groups faced with different behavioral costs (physical effort, cognitive effort, or delay) chose their preferred strategy to maximize rewards in foraging environments of varying richness. Environmental richness was dictated by the time it took to obtain a reward (the handling time) and the time between trials (the travel time). We hypothesized 1) that participants would approximate reward-maximizing behavior by preferring higher rewards and shorter delays, but that 2) participants confronted with effortful demands would be more likely to accept trials than those faced with passive delay, regardless of handling time and reward amount. The second hypothesis was based on the traditional idea that rewarded work should become appealing [@Eisenberger1992; @Inzlicht2018].

  We performed a mixed effects logistic regression to address the first hypothesis, pooling across the four conditions. Since participants in the effort groups were forced to travel when they failed to perform above threshold (see Methods), we distinguish between acceptances (not explicitly quitting a trial during the handling period) and completions (successfully obtaining a reward). Even so, the proportion of forced travels was low for cognitive effort (`r round(propFails[2, 2]$meanFT, digits = 2)` across participants) and 0 for the physical group. In line with our predictions, larger reward amounts significantly increased acceptance probabilities ($\beta$ = `r historyCoeffs_btw["Offer", "Estimate"]`, SE = `r historyCoeffs_btw["Offer", "Std. Error"]`, *p* `r report_p(historyCoeffs_btw["Offer", "Pr(>|z|)"])`), whereas longer delays decreased them ($\beta$ = `r historyCoeffs_btw["Handling", "Estimate"]`, SE = `r historyCoeffs_btw["Handling", "Std. Error"]`, *p* `r report_p(historyCoeffs_btw["Handling", "Pr(>|z|)"])`). Two additional regressors showed that having missed out on consecutive rewards decreased the probability of acceptance ($\beta$ = `r historyCoeffs_btw["recentTravels", "Estimate"]`, SE = `r historyCoeffs_btw["recentTravels", "Std. Error"]`, *p* `r report_p(historyCoeffs_btw["recentOffers", "Pr(>|z|)"])`), and that participants became more selective when recent offer history was richer ($\beta$ = `r historyCoeffs_btw["recentOffers", "Estimate"]`, SE = `r historyCoeffs_btw["recentOffers", "Std. Error"]`, *p* `r report_p(historyCoeffs_btw["recentOffers", "Pr(>|z|)"])`). The counterintuitive reduction in acceptances following consecutive unobtained rewards was driven by the wait group (removing this group from the model yielded a *p* > 0.1), likely because this group accepted the fewest offers. Comparing likelihood fits among lengths of offer history (e.g. sum over 2 or 3 past trials) suggested that participants were influenced by the `r historyIndex_btw + 1` most recent rewards (AIC = `r round(AIC(historySize_btw[[historyIndex_btw]]), digits = 2)`), which matched the offer presentation rotation (see Methods).

@@

# apply chi-squared
temp2 <- apply(temp, 1, function(row) {prop.test(as.numeric(row[3]), sum(as.numeric(row[3:4])), p = 0.5, conf.level = 0.95)})

@@

  The next set of analyses examined the hypothesis that both cognitive effort and physical effort groups would uniformly accept more trials than passive delay and easy groups. Fig. 2 (left) shows that the cognitive effort group consistently accepted more offers, which resulted in lower earnings (middle). One-way ANOVAs showed that differences among groups were significant both for overall proportion accepted (F(`r propAov[[1]]$Df[1]`, `r propAov[[1]]$Df[2]`) = `r round(propAov[[1]][1, 4], digits = 2)`, *p* `r report_p(propAov[[1]][1, 5])`), and total earnings (F(`r earnAov[[1]]$Df[1]`, `r earnAov[[1]]$Df[2]`) = `r round(earnAov[[1]][1, 4], digits = 2)`, *p* `r report_p(earnAov[[1]][1, 5])`). Post-hoc permutations (5000 iterations) comparing mean acceptance rates between every pair of costs confirmed that acceptance rates of the cognitive effort group (mean = `r filter(overallSumms, Cost == "Cognitive")$mProp`, SD = `r filter(overallSumms, Cost == "Cognitive")$sdProp`) were higher than those of the physical (mean = `r filter(overallSumms, Cost == "Physical")$mProp`, SD = `r filter(overallSumms, Cost == "Physical")$sdProp`; *p* `r report_p(filter(propTests, contrast == "Physical < Cognitive")$permPval)`, Cohen's D = `r filter(propTests, contrast == "Physical < Cognitive")$cohenD`) and wait groups (mean = `r filter(overallSumms, Cost == "Wait")$mProp`, SD = `r filter(overallSumms, Cost == "Wait")$sdProp`; *p* `r report_p(filter(propTests, contrast == "Cognitive > Wait")$permPval)`, Cohen's D = `r filter(propTests, contrast == "Cognitive > Wait")$cohenD`). It also showed that those faced with the easy task accepted more than those in the wait group, although responses from the easy group were more variable (mean Easy = `r filter(overallSumms, Cost == "Easy")$mProp`, SD = `r filter(overallSumms, Cost == "Easy")$sdProp`; *p* `r report_p(filter(propTests, contrast == "Wait < Easy")$permPval)`, Cohen's D = `r filter(propTests, contrast == "Wait < Easy")$cohenD`) (all other *p* > 0.05). The higher acceptance rates produced by the cognitive effort group resulted in significantly lower earnings (mean = `r filter(overallSumms, Cost == "Cognitive")$mEarn`, SD = `r filter(overallSumms, Cost == "Cognitive")$sdEarn`) than both physical (mean = `r filter(overallSumms, Cost == "Physical")$mEarn`, SD = `r filter(overallSumms, Cost == "Physical")$sdEarn`; *p* `r report_p(filter(earnTests, contrast == "Physical > Cognitive")$permPval)`, Cohen's D = `r filter(earnTests, contrast == "Physical > Cognitive")$cohenD`) and wait groups (mean = `r filter(overallSumms, Cost == "Wait")$mEarn`, SD = `r filter(overallSumms, Cost == "Wait")$sdEarn`; *p* `r report_p(filter(earnTests, contrast == "Cognitive < Wait")$permPval)`, Cohen's D = `r filter(earnTests, contrast == "Cognitive < Wait")$cohenD`). In addition, we found that the wait group earned more than the physical (*p* `r report_p(filter(earnTests, contrast == "Physical < Wait")$permPval)`, Cohen's D = `r filter(earnTests, contrast == "Physical < Wait")$cohenD`) and easy groups (mean Easy = `r filter(overallSumms, Cost == "Easy")$mEarn`, SD = `r filter(overallSumms, Cost == "Easy")$sdEarn`; *p* `r report_p(filter(earnTests, contrast == "Wait > Easy")$permPval)`, Cohen's D = `r filter(earnTests, contrast == "Wait > Easy")$cohenD`). Considering the low proportion of forced travels, performance was an unlikely explanation for the observed differences in choice.

  We next examined the optimality of these decisions. The foraging task was configured so that there was a single reward-maximizing strategy per block type (see Operationalization of Cost in Methods), and participants were warned that accepting everything would not maximize their gains. Accordingly, participants who over- or under-accepted earned the least (Fig. 2, right; general linear model with quadratic term, F = `r round(summary(lmEarn)$fstatistic[1], digits = 2)`, $\beta$ = `r round(summary(lmEarn)$coefficients[3,1], digits = 2)`, SE = `r round(summary(lmEarn)$coefficients[3, 2], digits = 2)`, $R^2$ = `r round(summary(lmEarn)$r.squared, digits = 2)`, *p* `r report_p(summary(lmEarn)$coefficients[3, 4])`). We performed group-level chi-squared tests of proportions to examine whether each cost biased decision makers to under- or over-accept 4 and 8 cent offers, which together should have been accepted 50% of the time. We found a significant over-acceptance of these trials by the cognitive effort group (proportion accepted = `r round(optimalProp_cost["Cognitive", "probability"], digits = 2)`; $\chi^2$ = `r round(optimalProp_cost["Cognitive", "chisquared"], digits = 2)`, *p* `r report_p(optimalProp_cost["Cognitive", "pval"])`), but under-acceptances by the physical (proportion accepted = `r round(optimalProp_cost["Physical", "probability"], digits = 2)`; $\chi^2$ = `r round(optimalProp_cost["Physical", "chisquared"], digits = 2)`, *p* `r report_p(optimalProp_cost["Physical", "pval"])`) and wait (proportion accepted = `r round(optimalProp_cost["Wait", "probability"], digits = 2)`; $\chi^2$ = `r round(optimalProp_cost["Wait", "chisquared"], digits = 2)`, *p* `r report_p(optimalProp_cost["Wait", "pval"])`) groups (no significant difference for the Easy group). Together, these findings show that even though participants did not always perform optimally, they responded appropriately to the statistics of the environment in each timing condition. Further, they provide evidence that cognitive effort can boost the value of an offer, as reflected by higher acceptances at the expense of earnings.

@@

  We next probed whether these results held uniformly across all parameter combinations. Fig. 3A shows the mean proportion of acceptances (± SEM) per combination of handling time, reward, and cost, with optimal acceptance rates depicted by the gray circles. Qualitatively, the figure confirms two important predictions. First, participants adapted to the richness of each timing block, gravitating towards optimality regardless of the cost they faced. Second, differences in cost were consistent across blocks, and acceptances were uniformly highest for the cognitive effort group. We tested these observations with a mixed-effects logistic regression, computing the probability of accepting an offer as a function of handling time, reward amount, and demand type (see Methods for details). As before, the model showed significant main effects of handling time (global intercept = `r round(fixef(mixLogis_main_btw$Wait)["(Intercept)"], digits = 2)`, $\beta$ = `r round(fixef(mixLogis_main_btw$Wait)["Handling"], digits = 2)`, SE = `r round(summary(mixLogis_main_btw$Wait)[[10]]["Handling", "Std. Error"], digits = 2)`, *p* `r report_p(summary(mixLogis_main_btw$Wait)[[10]]["Handling", "Pr(>|z|)"])`) and offer amount ($\beta$ = `r round(fixef(mixLogis_main_btw$Wait)["Offer"], digits = 2)`, SE = `r round(summary(mixLogis_main_btw$Wait)[[10]]["Offer", "Std. Error"], digits = 2)`, *p* `r report_p(summary(mixLogis_main_btw$Wait)[[10]]["Offer", "Pr(>|z|)"])`). Comparisons among all conditions are shown on Fig. 3B. The matrix of coefficients conveys that the cognitive effort group was significantly more likely to accept any offer than the physical ($\beta$ = `r round(mixSummary_btw$Betas["Physical", "Cognitive"], digits = 2)`, *p* `r report_p(mixSummary_btw$Pvals["Physical", "Cognitive"])`) and wait groups ($\beta$ = `r round(mixSummary_btw$Betas["Wait", "Cognitive"], digits = 2)`, *p* `r report_p(mixSummary_btw$Pvals["Wait", "Cognitive"])`), but not the easy group.

  These observations could not be more easily explained by simpler or more complex regression models. Fig. 3C shows AIC and BIC values for models of increasing complexity, starting with an intercept-only configuration (see Methods for details). Here we focus on AIC, as both metrics yielded comparable results. As predicted, the a priori model with all main effects performed better than those relying on a single parameter (a priori model pseudo $R^2$ = `r round(mixLogis_compare_btw$AllMainRsq, digits = 2)`), as well as the model with only main effects for handling time and reward amount (AIC~a-priori~ = `r filter(ME_evals$abicLogis_compare, Model == "AllMain")$AIC`, AIC~handling/reward~ = `r filter(ME_evals$abicLogis_compare, Model == "HR")$AIC`, $\chi^2$(`r ME_evals$HRvMain[[7]][2]`) = `r round(ME_evals$HRvMain$Chisq[2], digits = 2)`, *p* `r report_p(ME_evals$HRvMain[[8]][2])`).

  However, against our expectations, the a priori model was outperformed by models that added an interaction between handling and reward (AIC~HR_interaction~ = `r filter(ME_evals$abicLogis_compare, Model == "HR_interaction")$AIC`, $\chi^2$(`r ME_evals$MainvHRint[[7]][2]`) = `r round(ME_evals$MainvHRint$Chisq[2], digits = 2)`, *p* `r report_p(ME_evals$MainvHRint[[8]][2])`) as well as one that considered all possible interactions (AIC~all_interactions~ = `r filter(ME_evals$abicLogis_compare, Model == "Interaction")$AIC`, $\chi^2$(`r ME_evals$MainvAllint[[7]][2]`) = `r round(ME_evals$MainvAllint$Chisq[2], digits = 2)`, *p* `r report_p(ME_evals$MainvAllint[[8]][2])`). Even so, neither of these models offered further interpretable insights on the observed differences across costs, so we restrict our focus to the a priori model containing all main effects. These results further support the hypotheses that handling and reward amounts were similarly integrated across groups, and that the absolute rate of acceptances was affected by the demand faced. However, they offer partial support for the hypothesis that effort would induce higher acceptance rates, which was only true for those confronted with cognitive effort.

#### *Between-subjects: Consistency and Stability*

@@

  Next, we tested the hypothesis that foragers would exhibit within-trial consistency and across-trial stability in their choices. Participants were deemed consistent if they did not quit after engaging in a trial. To examine this, we plotted survival curves of trial quits across all participants (Fig. 4, top-left), with black crosses signaling either completed trials (at 2 s, 10 s, and 14 s) or forced travels. These curves show that most decisions to quit happened within the first second into the handling time, and participants rarely quit afterwards (percentage of quits after 1 s = `r  handlingQuits_btw$r * 100`). Physical and easy groups showed slightly lagged responses because the experiment allowed a one-second grace period for participants to begin gripping (marked by the dashed line). This resulted in some individuals opting to wait for that second to indicate an offer rejection. Plotting the quitting time distributions across blocks for the first second of the handling time (ECDFs on Fig. 4, top-right) showed that participants in the wait and cognitive groups responded faster as the experiment progressed. These response time observations suggested clarity in participant choices.

@@

  Stability was defined as a persistence in acceptance patterns throughout the experimental session. The scatter plot on Fig. 4 (bottom-left) shows the overall proportion accepted by each participant before and after the midpoint of the session. Paired permutations showed no differences in acceptances for wait and easy groups (*p* > 0.05), but we found significantly lower acceptances for physical (*p* `r report_p(prepost_btw$Perms[1])`, Cohen's D = `r round(prepost_btw$CohenD[1], digits = 2)`) and cognitive (*p* `r report_p(prepost_btw$Perms[2])`, Cohen's D = `r round(prepost_btw$CohenD[2], digits = 2)`) groups. Fig. 5 (left) shows that the reduction in acceptances in effort groups over time was steeper for longer handling times, a hallmark of effortful demands [@Treadway2009]. We then replicated the mixed-effects logistic regression from the previous section separately on each experimental half, which confirmed that differences among demands were stable (Fig. 5, right). Together, the consistency and stability displayed by participants suggest that decision makers had a clear representation of the environment.

>>>>

  There were three block types, in which handling times of 2, 10, or 14 seconds were paired with travel times of 14, 6, and 2 seconds, respectively. All combinations added up to 16 seconds, meaning that accepted trials took the same amount of time in all conditions, but rejected trials were shorter in environments with shorter travel times. Timing parameters were held constant within each 7-min block. Each block type was presented twice in pseudorandom order (total session length = 42 minutes). Reward amounts varied per trial (4, 8, or 20 cents with equal probability), with the constraint that every reward was presented twice every six trials. This prevented sequences from being dominated by a single amount during any window of time. Timing information was disclosed at the beginning of each block, and each trial's reward amount was displayed during a 2 s offer window before the trial began. Participants received training prior to the experimental session and were told about all possible timing conditions and reward amounts in order to avoid any biases that could arise through experience-dependent learning [@Dundon2020; @Garrett2020]. Finally, we encouraged participants to carefully evaluate their options by informing them that accepting all offers would not maximize their rewards.

@@

# Proportion of performance-based forced travels per subject and cost type
# move to 16.2.1.
propFails <- dataBtw %>% 
  filter(Cost != "Wait") %>% 
  group_by(SubjID, Cost) %>% 
  summarise(propFails = mean(rawChoice == 2), .groups = 'drop') %>%
  group_by(Cost) %>% 
  summarise(meanFT = mean(propFails),
            sdFT = sd(propFails), .groups = 'drop')

# overall acceptances across all conditions
allp <- dataBtw %>%
  group_by(SubjID) %>%
  summarise(p = mean(Choice)) %>%
  ungroup()
```

@@

  Four groups of participants faced with different behavioral demands (physical effort, cognitive effort, low-effort task engagement, or passive delay) chose their preferred strategy to maximize rewards in foraging environments of varying richness. Environmental richness was dictated by the time it took to obtain a reward (the handling time) and the time between trials (the travel time). We hypothesized 1) that participants would approximate reward-maximizing behavior by preferentially accepting higher rewards and shorter delays; 2) that participants confronted with effortful demands would be more likely to accept trials than those faced with passive delay, regardless of handling time and reward amount; and 3) that choices would be stable over time and consistent within trials.

  We performed a mixed-effects logistic regression to address the first hypothesis, pooling across the four conditions. Since participants in the effort groups were forced to travel when they failed to perform above threshold (see Methods), we distinguish between acceptances (not explicitly quitting a trial during the handling period) and completions (successfully obtaining a reward). Even so, the proportion of forced travels was low for the cognitive effort group (`r round(propFails[2, 2]$meanFT, digits = 2) * 100`% on average across participants) and 0 for the physical effort group (i.e. no forced travels in this group). Overall, participants accepted an average of `r round(mean(allp$p), digits = 2) * 100`% of trials (SD = `r round(sd(allp$p), digits = 2) * 100`%, range = `r round(min(allp$p), digits = 2) * 100`% to `r round(max(allp$p), digits = 2) * 100`%). In line with our predictions, larger reward amounts significantly increased acceptance probabilities ($\beta$ = `r historyCoeffs_btw["Offer", "Estimate"]`, SE = `r historyCoeffs_btw["Offer", "Std. Error"]`, *p* `r report_p(historyCoeffs_btw["Offer", "Pr(>|z|)"])`), whereas longer delays decreased them ($\beta$ = `r historyCoeffs_btw["Handling", "Estimate"]`, SE = `r historyCoeffs_btw["Handling", "Std. Error"]`, *p* `r report_p(historyCoeffs_btw["Handling", "Pr(>|z|)"])`). Two additional regressors showed that having missed out on consecutive rewards decreased the probability of acceptance ($\beta$ = `r historyCoeffs_btw["recentTravels", "Estimate"]`, SE = `r historyCoeffs_btw["recentTravels", "Std. Error"]`, *p* `r report_p(historyCoeffs_btw["recentOffers", "Pr(>|z|)"])`), and that participants became more selective when recent offer history was richer ($\beta$ = `r historyCoeffs_btw["recentOffers", "Estimate"]`, SE = `r historyCoeffs_btw["recentOffers", "Std. Error"]`, *p* `r report_p(historyCoeffs_btw["recentOffers", "Pr(>|z|)"])`). The counterintuitive reduction in acceptances following consecutive unobtained rewards was driven by the Wait group (removing this group from the model yielded a *p* > 0.1).

@@

# apply chi-squared test of proportion per group (pooled trials across individuals)
temp2 <- apply(temp, 1, function(row) {prop.test(as.numeric(row[3]), sum(as.numeric(row[3:4])), p = 0.5, conf.level = 0.95)})

@@

  The next set of analyses examined the hypothesis that both cognitive effort and physical effort groups would uniformly accept more trials than passive delay and easy groups. Fig. 2 (left) shows that the cognitive effort group consistently accepted more offers, which resulted in lower earnings (middle). One-way ANOVAs showed that differences among groups were significant both for overall proportion accepted (F(`r propAov[[1]]$Df[1]`, `r propAov[[1]]$Df[2]`) = `r round(propAov[[1]][1, 4], digits = 2)`, *p* `r report_p(propAov[[1]][1, 5])`), and total earnings (F(`r earnAov[[1]]$Df[1]`, `r earnAov[[1]]$Df[2]`) = `r round(earnAov[[1]][1, 4], digits = 2)`, *p* `r report_p(earnAov[[1]][1, 5])`). Post-hoc permutations (5000 iterations) comparing mean acceptance rates between every pair of costs confirmed that acceptance rates of the cognitive effort group (mean = `r filter(overallSumms, Cost == "Cognitive")$mProp`, SD = `r filter(overallSumms, Cost == "Cognitive")$sdProp`) were higher than those of the physical (mean = `r filter(overallSumms, Cost == "Physical")$mProp`, SD = `r filter(overallSumms, Cost == "Physical")$sdProp`; *p* `r report_p(filter(propTests, contrast == "Physical < Cognitive")$permPval)`, Cohen's D = `r filter(propTests, contrast == "Physical < Cognitive")$cohenD`) and wait groups (mean = `r filter(overallSumms, Cost == "Wait")$mProp`, SD = `r filter(overallSumms, Cost == "Wait")$sdProp`; *p* `r report_p(filter(propTests, contrast == "Cognitive > Wait")$permPval)`, Cohen's D = `r filter(propTests, contrast == "Cognitive > Wait")$cohenD`). It also showed that those faced with the easy task accepted more than those in the wait group, although responses from the easy group were more variable (mean Easy = `r filter(overallSumms, Cost == "Easy")$mProp`, SD = `r filter(overallSumms, Cost == "Easy")$sdProp`; *p* `r report_p(filter(propTests, contrast == "Wait < Easy")$permPval)`, Cohen's D = `r filter(propTests, contrast == "Wait < Easy")$cohenD`) (all other *p* > 0.05). In line with its higher acceptance rates, the cognitive effort group had significantly lower earnings (mean = `r filter(overallSumms, Cost == "Cognitive")$mEarn`, SD = `r filter(overallSumms, Cost == "Cognitive")$sdEarn`) than either the physical group (mean = `r filter(overallSumms, Cost == "Physical")$mEarn`, SD = `r filter(overallSumms, Cost == "Physical")$sdEarn`; *p* `r report_p(filter(earnTests, contrast == "Physical > Cognitive")$permPval)`, Cohen's D = `r filter(earnTests, contrast == "Physical > Cognitive")$cohenD`) or the wait group (mean = `r filter(overallSumms, Cost == "Wait")$mEarn`, SD = `r filter(overallSumms, Cost == "Wait")$sdEarn`; *p* `r report_p(filter(earnTests, contrast == "Cognitive < Wait")$permPval)`, Cohen's D = `r filter(earnTests, contrast == "Cognitive < Wait")$cohenD`). In addition, the wait group earned more than the physical (*p* `r report_p(filter(earnTests, contrast == "Physical < Wait")$permPval)`, Cohen's D = `r filter(earnTests, contrast == "Physical < Wait")$cohenD`) and easy groups (mean Easy = `r filter(overallSumms, Cost == "Easy")$mEarn`, SD = `r filter(overallSumms, Cost == "Easy")$sdEarn`; *p* `r report_p(filter(earnTests, contrast == "Wait > Easy")$permPval)`, Cohen's D = `r filter(earnTests, contrast == "Wait > Easy")$cohenD`).

  We next examined the optimality of these decisions. The foraging task was configured so that there was a single reward-maximizing strategy per block type, and participants were informed that accepting everything would not maximize their gains. Accordingly, participants who over- or under-accepted earned the least (Fig. 2, right; general linear model with quadratic term, F = `r round(summary(lmEarn)$fstatistic[1], digits = 2)`, $\beta$ = `r round(summary(lmEarn)$coefficients[3,1], digits = 2)`, SE = `r round(summary(lmEarn)$coefficients[3, 2], digits = 2)`, $R^2$ = `r round(summary(lmEarn)$r.squared, digits = 2)`, *p* `r report_p(summary(lmEarn)$coefficients[3, 4])`). We performed group-level chi-squared tests of proportions (with trials pooled across participants) to examine whether each cost biased decision makers to under- or over-accept 4 and 8 cent offers, which collectively should have been accepted 50% of the time across the three timing environments. We found a significant over-acceptance of these trials by the cognitive effort group (proportion accepted = `r round(optimalProp_cost["Cognitive", "probability"], digits = 2)`; $\chi^2$ = `r round(optimalProp_cost["Cognitive", "chisquared"], digits = 2)`, *p* `r report_p(optimalProp_cost["Cognitive", "pval"])`), but under-acceptances by the physical effort group (proportion accepted = `r round(optimalProp_cost["Physical", "probability"], digits = 2)`; $\chi^2$ = `r round(optimalProp_cost["Physical", "chisquared"], digits = 2)`, *p* `r report_p(optimalProp_cost["Physical", "pval"])`) and wait group (proportion accepted = `r round(optimalProp_cost["Wait", "probability"], digits = 2)`; $\chi^2$ = `r round(optimalProp_cost["Wait", "chisquared"], digits = 2)`, *p* `r report_p(optimalProp_cost["Wait", "pval"])`), with no significant difference for the Easy group. Together, the findings discussed so far indicate that although participants reacted appropriately to reward and timing contingencies, departures from optimality were systematically influenced by what form of cost was at stake. The findings raise the possibility that cognitive effort can boost the value of an offer, as reflected by higher acceptances at the expense of earnings.

@@

  We next probed whether these results held uniformly across reward and timing parameter combinations. Fig. 3A shows the mean proportion of acceptances (± SEM) per combination of handling time, reward, and cost type, with reward-maximizing acceptance rates depicted by the gray circles. Qualitatively, the figure confirms two important predictions. First, participants adapted to the richness of each timing block, gravitating towards optimality regardless of the cost they faced. Second, effects of cost type were consistent across the three timing conditions, and acceptances were uniformly highest for the cognitive effort group. We tested these observations with a mixed-effects logistic regression model, which estimated the probability of accepting an offer as a function of handling time, reward amount, and cost type (see Methods for details). The model showed significant main effects of handling time ($\beta$ = `r round(fixef(mixLogis_main_btw$Wait)["Handling"], digits = 2)`, SE = `r round(summary(mixLogis_main_btw$Wait)[[10]]["Handling", "Std. Error"], digits = 2)`, *p* `r report_p(summary(mixLogis_main_btw$Wait)[[10]]["Handling", "Pr(>|z|)"])`) and offer amount ($\beta$ = `r round(fixef(mixLogis_main_btw$Wait)["Offer"], digits = 2)`, SE = `r round(summary(mixLogis_main_btw$Wait)[[10]]["Offer", "Std. Error"], digits = 2)`, *p* `r report_p(summary(mixLogis_main_btw$Wait)[[10]]["Offer", "Pr(>|z|)"])`). Comparisons among all conditions are shown on Fig. 3B. The cognitive effort group was significantly more likely to accept offers than the physical effort group ($\beta$ = `r round(mixSummary_btw$Betas["Physical", "Cognitive"], digits = 2)`, *p* `r report_p(mixSummary_btw$Pvals["Physical", "Cognitive"])`) or the wait groups ($\beta$ = `r round(mixSummary_btw$Betas["Wait", "Cognitive"], digits = 2)`, *p* `r report_p(mixSummary_btw$Pvals["Wait", "Cognitive"])`), but not the easy group.

  These observations were not better explained by simpler or more complex regression models. Fig. 3C shows AIC and BIC values for models of increasing complexity, starting with an intercept-only configuration (see Methods for details). Here we focus on AIC, as both metrics yielded comparable results. As predicted, the a priori model with all main effects performed better than those relying on a single parameter (a priori model pseudo $R^2$ = `r round(mixLogis_compare_btw$AllMainRsq, digits = 2)`), as well as the model with only main effects for handling time and reward amount (AIC~a-priori~ = `r filter(ME_evals$abicLogis_compare, Model == "AllMain")$AIC`, AIC~handling/reward~ = `r filter(ME_evals$abicLogis_compare, Model == "HR")$AIC`, $\chi^2$(`r ME_evals$HRvMain[[7]][2]`) = `r round(ME_evals$HRvMain$Chisq[2], digits = 2)`, *p* `r report_p(ME_evals$HRvMain[[8]][2])`).

  Contrary to our expectations, the a priori model was outperformed by models that added an interaction between handling and reward (AIC~HR_interaction~ = `r filter(ME_evals$abicLogis_compare, Model == "HR_interaction")$AIC`, $\chi^2$(`r ME_evals$MainvHRint[[7]][2]`) = `r round(ME_evals$MainvHRint$Chisq[2], digits = 2)`, *p* `r report_p(ME_evals$MainvHRint[[8]][2])`) as well as one that considered all possible interactions (AIC~all_interactions~ = `r filter(ME_evals$abicLogis_compare, Model == "Interaction")$AIC`, $\chi^2$(`r ME_evals$MainvAllint[[7]][2]`) = `r round(ME_evals$MainvAllint$Chisq[2], digits = 2)`, *p* `r report_p(ME_evals$MainvAllint[[8]][2])`). However, upon examination, neither of these models offered further interpretable insights on the observed differences across costs. The overall pattern of results supported the hypotheses that handling and reward amounts were integrated in the expected fashion, and that the rate of acceptances was affected by the demand faced. The results offered only partial support for the hypothesis that effort would induce higher acceptance rates; higher acceptance rates were observed for cognitive effort but not for physical effort.

#### *Consistency and Stability of Choices*

@@

  Next, we tested the hypothesis that foragers would exhibit within-trial consistency and across-trial stability in their choices. Participants were deemed consistent if they tended not to quit after engaging in a trial (that is, if they pressed the spacebar to reject a trial either quickly or not at all). To examine this, we plotted survival curves of trial rejection times across all participants (Fig. 4, top-left), with black crosses signaling censored observations either for completed trials (at 2 s, 10 s, and 14 s) or forced travels. These curves show that most decisions to quit happened within the first second into the handling time, and participants rarely quit afterwards (only `r  handlingQuits_btw$r * 100`% of quits occurred later than 1 s). The pattern of results in the Physical and Easy groups reflected that participants sometimes quit by letting the 1 s initial grace period expire without gripping (rather than by pressing the spacebar); such trials were coded as quits rather than as forced travels. Plotting the quitting time distributions across blocks for the first 1 s of the handling time (ECDFs on Fig. 4, top-right) showed that participants in the wait and cognitive groups responded faster as the experiment progressed. These response time observations suggested clarity in participant choices.

@@

  Stability was defined as the maintenance of similar acceptance rates throughout the experimental session. The scatterplot in Fig. 4 (bottom-left and bottom-right) shows the overall proportion accepted by each participant before and after the midpoint of the session. Paired permutations showed no change in acceptance rates for the wait and easy groups (*p* > 0.05), but we found significantly lower acceptance rates in the second half of the session for the physical effort group (*p* `r report_p(prepost_btw$Perms[1])`, Cohen's D = `r round(prepost_btw$CohenD[1], digits = 2)`) and the cognitive effort group (*p* `r report_p(prepost_btw$Perms[2])`, Cohen's D = `r round(prepost_btw$CohenD[2], digits = 2)`). Fig. 5 (left) shows that the reduction in acceptance rates in effort groups over time was steeper for longer handling times, a hallmark of effortful demands [@Treadway2009]. We then replicated the mixed-effects logistic regression from the previous section separately for each half of the session, which confirmed that differences among demands were stable (Fig. 5, right). Together, the consistency and stability of behavior suggested that decision makers had a clear representation of the environment.
